---
title: 'Basic Data Visualization in R: Hands on Practice'
author: "Holly, Rebecca & Alicia Vallorani"
date: "`r Sys.time()`"
output:
  html_document:
    code_foldering: show
    mathjax: default
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# handy package manager that installs and loads packages used in this document
if (!("pacman" %in% installed.packages()[,"Package"])) { install.packages("pacman") } 
pacman::p_load(tidyverse, psych, GGally, Rmisc, interactions) # install (if needed) and load packages
```

# Hands on practice: data analyses
For this hands on practice, load the `spi` data set from the `psych` package.

1. Check the help file, structure, and first few observations of the data and the structure of the data.
```{r}
# Load data and convert sex and education to factors
source("R/load_spi.R")

# Write your code here!

```

2. Examine the distribution of your continuous variables.
```{r}

# Write your code here!

```

3. Create a scatter plot that examines the relation between age and health.
```{r}

# Write your code here!

```

4. Add a categorical variable to your scatterplot such as sex or exercise using color and/or shape as the asthetic. Try playing with the alpha level of the points. 
```{r}

# Write your code here!

```

5. Add a regression line to your scatterplot
```{r}

# Write your code here

```

6. Try using ggpairs to examine multiple relations between your variables simultaneously. You can examine the entire dataset or select particular variables to include using dplyr::select() in the pipe.
```{r}

# Write your code here!

```


7. Use a bar graph to examine a count of ER visits.
```{r}

# Write your code here!

```


8. Use a bar graph to examine mean health by ER visits. You can use na.rm = TRUE inside geom_bar() to remove missing data.
```{r}

# Write your code here!

```

9. Create a bar graph examining health by ER visits with error bars using Rmisc and summarySE().
```{r}

# Write your code here!

```

10. Use a bar graph to examine a three way interaction between health, ER and sex. Hint: Use color to define sex.
```{r}

# Write your code here!

```

11. Say you were interested in relations between health, wellness and sex. How would you visualize this relation?
```{r}

# Write your code here!

```

12. Try exporting one of your favorite plots you made!
```{r}

# Write your code here!

```


