---
title: "Basic Analyses in R: Hands on Practice"
author: "Daniel Albohn, Kayla Brown, & Yiming Qian"
date: "`r Sys.time()`"
output:
  html_document:
    code_foldering: show
    mathjax: default
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: show
---

```{r packages, message = FALSE}
pkg_list <- c("tidyverse", "psych", "ez", "rcompanion", "knitr", "car",
              "afex", "ggfortify", "Hmisc", "emmeans", "jtools", "apaTables")
purrr::walk(pkg_list, library, quietly = TRUE, character.only = TRUE)
```

# Hands on practice: data analyses
For this hands on practice, load the `spi` data set from the `psych` package.

1. Check the help file, structure, and first few observations of the data

```{r}
# Write your code here!

```

2. Change the integer categorical variables to factors by cross-referencing
the help page and filling in the pipeline below (note when knitting this
document, you may want to change `eval` to `TRUE` once you complete this
exercise)

```{r eval = FALSE}
spi <- spi %>%
  mutate(.,
    sex_fac = case_when(),
    health_fac = case_when(),
    education_fac = case_when(),
    smoke_fac = case_when(),
    exer_fac = case_when()
  )

# Write your code here!

```

## Descriptives and Chi-squared

1. Use either `describe()` or `skim()` to describe the data by each categorical
variable

```{r}
# Write your code here!

```


2. Describe the data by both categorical variables at once

```{r}
# Write your code here!

```


3. Perform a chi-squared test assessing whether there is a differences in smoking
by education level

```{r}
# Write your code here!

```

## Correlation

1. Get the covariance and correlation matrix for the relationship between
three numeric values of your choosing

```{r}
# Write your code here!

```

2. Perform a test of significance on two of the variables selected above

```{r}
  # Write your code here!

```

## Linear Models

## ANOVAs